Survival Analysis

### 2014-04-02 presentation.Rnw
Fix stuff
Add dataset graphs and tweak equations

<<echo=FALSE>>=
library(ggplot2)
library(survival)
library(KMsurv) # survival data sets
library(GGally) # survival curve support for ggplot2

@@

\begin{frame}{Example}
\begin{center}
<<bfeed-survival, echo=FALSE, fig=TRUE, height=4>>=
bfeed.fit <- survfit(Surv(duration, delta) ~ 1, data = bfeed.short)
ggsurv(bfeed.fit) +
  ggtitle('Breast Feeding Duration') +
  xlab('Duration in Weeks') +
  ylab('Survival') +
  geom_vline(aes(xintercept = 27),
    colour = 'red',
    linetype = 'longdash')
@
\end{center}
\end{frame}

\begin{frame}{Example}
\begin{center}
<<bfeed-hazard, echo=FALSE, fig=TRUE, height=4>>=
# bfeed.haz <- basehaz(coxph(Surv(duration, delta) ~ 1, data = bfeed.short))
bfeed.haz <- with(bfeed.fit, -log(data.frame(
  hazard = surv, # haz = -log(surv)
  upper = upper,
  lower = lower)))
bfeed.haz$week <- as.numeric(rownames(bfeed.haz))

@@

\begin{frame}{Estimators}
Kaplan-Meier Max Likelihood Estimator for $\hat S(t)$

\vspace{0.2in}
\begin{tabular}{ll}
  $t_i$ & time interval \\
  $d_i$ & \# events in $t_i$ \\
  $n_i$ & \# subjects \emph{at risk} in $t_i$ \\
\end{tabular}

\begin{eqnarray*}
  \hat S(t_0) & = & 1 \\
  \hat S(t)   & = & \prod_{t_i < t} \frac{n_i - d_i}{n_i} \\
              & = & \frac{n_i - d_i}{n_i} \cdot S(t_{i - 1}) \\ \\
    n_{i + 1} & = & n_i - d_i - c_i
\end{eqnarray*}
\end{frame}

\begin{frame}{Models}
(Cox) Proportional Hazards Model

@@

Accelerated Failure Time Model

>>>>

<<echo=FALSE>>=
library(ggplot2)
library(gridExtra) # multi-plot graphs
library(survival)
library(KMsurv) # survival data sets
library(GGally) # survival curve support for ggplot2
library(xtable) # make latex tables

@@

\begin{frame}{Dataset}
\begin{itemize}
  \item National Labor Survey of Youth
  \item Examined pregnancies before 1983
  \item Restricted to first-born child
  \item Restricted to gestation of between 20 and 45 weeks
  \item Restricted to subjects with complete responses
\end{itemize}

\tiny
<<bfeed-table, echo = FALSE, results = TEX>>=
bfeed.ids <- sample(rownames(bfeed.short), size = 8)
bfeed.table <- bfeed.short[bfeed.ids, ]

print(xtable(bfeed.table))
@
\end{frame}

\begin{frame}{Dataset}
\begin{center}
<<bfeed-survival, echo=FALSE, fig=TRUE, height=4>>=
bfeed.fit <- survfit(Surv(duration, delta) ~ 1, data = bfeed.short)
ggsurv(bfeed.fit) +
  ggtitle('Breast Feeding Duration') +
  xlab('Duration in Weeks') +
  ylab('Survival') +
  geom_vline(aes(xintercept = 27),
    colour = 'red',
    linetype = 'longdash')
@
\end{center}
\end{frame}

\begin{frame}{Dataset}
\begin{center}
<<bfeed-hazard, echo=FALSE, fig=TRUE, height=4>>=
# bfeed.haz <- basehaz(coxph(Surv(duration, delta) ~ 1, data = bfeed.short))
bfeed.haz <- with(bfeed.fit, -log(data.frame(
  hazard = surv, # haz = -log(surv)
  upper = upper,
  lower = lower)))
bfeed.haz$week <- as.numeric(rownames(bfeed.haz))

@@

\begin{frame}{Dataset}
\begin{center}
<<bfeed-by-education, echo = FALSE, fig = TRUE, height = 4>>=
bfeed.short$college <- bfeed.short$yschool > 12
bfeed.fit <- survfit(Surv(duration, delta) ~ college, data = bfeed.short)

ggsurv(bfeed.fit) +
  ggtitle('Breast Feeding Duration by Education') +
  xlab('Duration in Weeks') +
  ylab('Survival') +
  geom_vline(aes(xintercept = 27),
    colour = 'red',
    linetype = 'longdash')
@
\end{center}
\end{frame}

\begin{frame}{Dataset}
\begin{center}
<<bfeed-by-poverty, echo = FALSE, fig = TRUE, height = 5>>=
bfeed.fit <- survfit(Surv(duration, delta) ~ poverty, data = bfeed.short)

bfeed.poverty.short <- ggsurv(bfeed.fit) +
  ggtitle('Breast Feeding Duration by Poverty, Duration < 30 Weeks') +
  xlab('Duration in Weeks') +
  ylab('Survival') +
  geom_vline(aes(xintercept = 27),
    colour = 'red',
    linetype = 'longdash')

bfeed.fit <- survfit(Surv(duration, delta) ~ poverty, data = subset(bfeed, duration > 20))

bfeed.poverty.all <- ggsurv(bfeed.fit) +
  ggtitle('Breast Feeding Duration by Poverty, Duration > 20 Weeks') +
  xlab('Duration in Weeks') +
  ylab('Survival') +
  geom_vline(aes(xintercept = 27),
    colour = 'red',
    linetype = 'longdash')

grid.arrange(bfeed.poverty.short, bfeed.poverty.all, nrow = 2)
@
\end{center}
\end{frame}

\begin{frame}{Kaplan-Meier Max Likelihood Estimator for $\hat S(t)$}
\begin{tabular}{ll}
  $t_i$ & time interval \\
  $d_i$ & \# events in $t_i$ \\
  $n_i$ & \# subjects \emph{at risk} in $t_i$ \\
\end{tabular}

\begin{eqnarray*}
                             n_{i + 1} & = & n_i - d_i - c_i \\
                             \hat S(0) & = & 1 \\
                             \hat S(t) & = & \prod_{t_i = 0}^{t} \frac{n_i - d_i}{n_i} \\
                                       & = & \frac{n_i - d_i}{n_i} \cdot S(t_{i - 1}) \\
  \mathrm{Var}\left[ \hat S(t) \right] & = & \sum_{t_i = 0}^{t} \frac{d_i}{n_i (n_i - d_i)}
\end{eqnarray*}
\end{frame}

\begin{frame}{Models}
(Cox) Proportional Hazards Model
% semi-parametric
% hazards function given covariates (factors)
% H_0: all \Beta_i = 0

@@

Accelerated Failure Time Model
% parametric, specifies log(T)
%   exponential
%   weibull
%   lognormal
%   log-log
%   gamma
